---
title: "t2d-samafs phenotype distribution"
output:
  pdf_document:
    keep_tex: true
header-includes:
  - \usepackage[margin=1in]{geometry}
date: "2025-01-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Available traits:

* cystatin_c - log transform
* a2h_ins - log transform
* creatinine - log transform
* adiponectin - log transform
* leptin - log transform
* chol -- log transform
* ldl -- log transform
* hdl -- log transform
* tg -- log transform
* bmi -- log transform
* hipc -- log transform
* wasitc -- log transform
* whr -- log transform
* dbp -- log transform
* height 
* fast_glu -- log transform
* sbp -- log transform
* fast_ins -- log transform
* a2h_glu -- log transform
* weight -- log transform


```{r}
# load phenotype file
pheno = read.table("/hpc/dctrl/tt207/meta_analysis_aim/samafs/phs000462.v2.pht002844.v2.p1.c1.T2D_GENES_SAMAFS_Project2_WGS_Subject_Phenotypes.DS-DIAB-IRB-RD.txt.gz",
                   header = TRUE) 
```

```{r fig.width=7, fig.height=3.5}
# traits of interest
traits <- c("height", "fast_glu", "sbp", "fast_ins", "a2h_glu", "a2h_ins", 
            "creatinine", "adiponectin", "leptin", "chol", "ldl", 
            "hdl", "tg", "bmi", "hipc", "waistc", "whr", "dbp", "weight")


for (i in seq_along(traits)) {
  current_trait <- traits[i]
  print(current_trait)
  
  # fit regression model (untransformed)
  par(mfrow = c(1, 2))
  obj <- lm(pheno[[current_trait]] ~ pheno$sex + pheno$age)
  y2 <- obj$residuals
  hist(y2, main = paste("Residuals for", current_trait), 
       xlab = paste("Residuals ", current_trait))
  qqnorm(y2, main = paste("Residuals for", current_trait))
  qqline(y2, col = "blue")
  
  # fit regression model (apply log transformation)
  obj <- lm(log(pheno[[current_trait]]) ~ pheno$sex + pheno$age)
  y2 <- obj$residuals
  hist(y2, main = paste("Residuals for \n Log transformed", current_trait), 
       xlab = paste("Residuals log", current_trait))
  qqnorm(y2, main = paste("Residuals for \n Log transformed", current_trait))
  qqline(y2, col = "blue")
}
```

```{r fig.width=7, fig.height=3.5}
cleaned_pheno <- pheno %>%
  filter(!is.na(cystatin_c), !is.nan(cystatin_c), !is.infinite(cystatin_c)) %>%
  filter(!is.na(sex), !is.na(age)) 

par(mfrow = c(1, 2))
obj <- lm(as.numeric(cystatin_c) ~ sex + age, data = cleaned_pheno)
y2 <- obj$residuals
hist(y2, main = "Residuals for cystatin_c", 
     xlab = "Residuals cystatin_c")
qqnorm(y2, main = "Residuals for cystatin_c")
qqline(y2, col = "blue")

obj <- lm(log(as.numeric(cystatin_c)) ~ sex + age, data = cleaned_pheno)
y2 <- obj$residuals
hist(y2, main = "Residuals for log cystatin_c", 
     xlab = "Residuals log cystatin_c")
qqnorm(y2, main = "Residuals for log cystatin_c")
qqline(y2, col = "blue")

```



