```{r}
library(tidyverse)
library(genio)
```

# write id's for sex (not available in fam)
```{r}
pheno = read.table("phs000462.v2.pht002844.v2.p1.c1.T2D_GENES_SAMAFS_Project2_WGS_Subject_Phenotypes.DS-DIAB-IRB-RD.txt.gz", header = TRUE)

table(pheno$sex)
male = pheno %>% filter(sex == 1) %>% mutate(famid = SUBJECT_ID, id = SUBJECT_ID) %>% select(famid, id)
female = pheno %>% filter(sex == 2) %>% mutate(famid = SUBJECT_ID, id = SUBJECT_ID) %>% select(famid, id)
# write group id's
write.table(male, "ids_male.txt", col.names = TRUE, 
            row.names = FALSE, quote = FALSE)
write.table(female, "ids_female.txt", col.names = TRUE, 
            row.names = FALSE, quote = FALSE)
```

# write covar
```{r}
grm <- read_grm( 'exome_chip_qc' )
eigenvec <- read_eigenvec( 'exome_chip_qc' )

pheno <- pheno[ match( grm$fam$id, pheno$SUBJECT_ID ), ]
# now that data is aligned, include all PCs as a single, convenient covariate
pheno$PCs <- eigenvec$eigenvec

```

set of traits 1:
* cystatin_c - log transform
* a2h_ins - log transform
* creatinine - log transform
* adiponectin - log transform
* leptin - log transform


```{r}
pheno_1 = pheno %>% mutate(LOG_cystatin_c = log(as.numeric(cystatin_c)),
                           LOG_a2h_ins = log(a2h_ins),
                           LOG_creatinine = log(creatinine),
                           LOG_adiponectin = log(adiponectin),
                           LOG_leptin = log(leptin))
```

set of traits 2:
* tg -- log transform
* bmi
* wasitc 
* whr 

```{r}
pheno_2 = pheno %>% mutate(LOG_tg = log(tg))
```

set of traits 3:
* hipc -- log transform
* dbp -- log transform
* chol -- log transform
* ldl -- log transform
* hdl -- log transform
```{r}
pheno_3 = pheno %>% mutate(LOG_hipc = log(hipc),
                           LOG_dbp = log(dbp),
                           LOG_chol = log(chol),
                           LOG_ldl = log(ldl),
                           LOG_hdl = log(hdl))
```

set of traits 4:
* bmi -- log transform
* whr -- log transform
* waistc -- log transform
```{r}
pheno_4 = pheno %>% mutate(LOG_bmi = log(bmi),
                           LOG_whr = log(whr),
                           LOG_waistc = log(waistc))
```

```{r}
pheno_5 = pheno %>% mutate(LOG_fast_glu = log(fast_glu),
                           LOG_fast_ins = log(fast_ins),
                           LOG_sbp = log(sbp),
                           LOG_a2h_glu = log(a2h_glu),
                           LOG_weight = log(weight))
```


```{r}
traits = c("cystatin_c", "a2h_ins", "creatinine", "adiponectin", "leptin", "chol", "ldl", "hdl", "tg", "bmi", "hipc",
           "waistc", "whr", "dbp")
trait_age = c("CYSTATIN_C_AGE", "A2H_INS_AGE", "CREATININE_AGE", "ADIPONECTIN_AGE", "LEPTIN_AGE", 
              "CHOL_AGE", "LDL_AGE", "HDL_AGE", "TG_AGE", "BMI_AGE", "HIPC_AGE",
           "WAISTC_AGE", "WHR_AGE", "DBP_AGE")


traits = c("LOG_cystatin_c", "LOG_a2h_ins", "LOG_creatinine", "LOG_adiponectin", "LOG_leptin")
trait_age = c("CYSTATIN_C_AGE", "A2H_INS_AGE", "CREATININE_AGE", "ADIPONECTIN_AGE", "LEPTIN_AGE")


traits = c("LOG_hipc", "LOG_dbp", "LOG_chol", "LOG_ldl", "LOG_hdl")
trait_age = c("HIPC_AGE", "DBP_AGE", "CHOL_AGE", "LDL_AGE", "HDL_AGE")


traits = c("LOG_bmi", "LOG_whr", "LOG_waistc")
trait_age = c("BMI_AGE", "WHR_AGE", "WAISTC_AGE")

traits = c("LOG_fast_glu", "LOG_fast_ins", "LOG_sbp", "LOG_a2h_glu", "LOG_weight")
trait_age = c("FAST_GLU_AGE", "FAST_INS_AGE", "SBP_AGE", "A2H_GLU_AGE", "WEIGHT_AGE")

for (i in seq_along(traits)) {
  current_trait <- traits[i]
  current_trait_age <- trait_age[i]
  
  # Create the covar dataframe with the current trait and trait_age
  covar <- pheno_5 %>%
    select(SUBJECT_ID, sex, all_of(current_trait), all_of(current_trait_age), PCs) 
  
 
  write.table(covar, paste0("covar_", current_trait, "_trait_age.txt"), col.names = TRUE, 
            row.names = FALSE, quote = FALSE)
}




colnames(pheno)
table(pheno$t2d)
hist(pheno$cystatin_c)
sum(is.na(pheno$cystatin_c))
sum(is.na(pheno$a2h_glu)) # 151 missing
sum(is.na(pheno$a2h_ins)) # 210 missing
sum(is.na(pheno$cystatin_c)) # 43 missing
sum(is.na(pheno$creatinine)) # 329 missing
sum(is.na(pheno$adiponectin)) # 280
sum(is.na(pheno$lepton)) 
sum(is.na(pheno$chol)) # 19
sum(is.na(pheno$ldl)) # 31
sum(is.na(pheno$hdl)) # 19
sum(is.na(pheno$tg)) 
```

