---
title: "hchs-sol phenotype distribution"
output:
  pdf_document:
    keep_tex: true
header-includes:
  - \usepackage[margin=1in]{geometry}
date: "2025-01-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Available traits:

* HEIGHT
* BMI -- log transformed
* WAIST_HIP -- log transformed
* INSULIN_FAST -- log transformed
* LABA70: glucose fasting (mg/dL) -- log transformed
* LABA76: glucose post OGTT (mg/dL) -- log transformed
* LABA101: cystatin C (mg/L) -- log transformed
* LABA91: High-sensitivity C-Reactive Protein (mg/L) -- log transformed
* LABA66: Total cholesterol (mg/dL) -- log transformed
* LABA67: Triglycerides (mg/dL) -- log transformed
* LABA68: HDL-cholesterol (mg/dL) -- log transformed
* LABA69: LDL-cholesterol (mg/dL) -- log transformed
* ANTA4: Weight (kg) -- log transformed
* ANTA10A: Waist Girth (cm) -- log transformed
* ANTA10B: Hip Girth (cm) -- log transformed
* SBPA5: Average Systolic -- log transformed
* SBPA6: Average Diastolic -- log transformed
* INSULIN_OGTT: OGTT -- log transformed
* LABA1: White Blood Count -- log transformed
* LABA10: % Neutrophils 
* LABA11: % Lymphocytes 
* LABA12: % Monocytes 
* LABA13: % Eosiniphils 
* LABA14: % Basophils 
* LABA2: Red Blood Count 
* LABA3: Hemoglobin -- log transformed
* LABA9: Platelet Count -- log transformed
* LABA74: ALT -- log transformed
* LABA75: AST -- log transformed
* LABA102: GGT -- log transformed
* LABA103: Ferritin -- log transformed
* LABA82: Iron -- log transformed
* SLPA54: AHI -- log transformed
  
\newpage

```{r}
# load phenotype file
pheno = read.csv("/hpc/group/ochoalab/zh105/project2/hchs-sol/Ia/heritability/hchs_phen.csv", 
                 header = TRUE) %>% rename(sex = GENDERNUM, age = AGE)
```

```{r fig.width=7, fig.height=3.5}
# traits of interest
traits = c("HEIGHT", "BMI", "WAIST_HIP", 
           "INSULIN_FAST", "LABA70", "LABA76", "LABA101", 
           "LABA91", "LABA66", "LABA67", "LABA68", "LABA69",
           "ANTA4", "ANTA10A", "ANTA10B", "SBPA5", "SBPA6",
           'INSULIN_OGTT', 'LABA1', 'LABA11',
           'LABA2', 'LABA3', 'LABA9',
           'LABA74', 'LABA75',  'LABA102', 'LABA103', 'LABA82')

for (i in seq_along(traits)) {
  current_trait <- traits[i]
  print(current_trait)
  
  # fit regression model (untransformed)
  par(mfrow = c(1, 2))
  obj <- lm(pheno[[current_trait]] ~ pheno$sex + pheno$age)
  y2 <- obj$residuals
  hist(y2, main = paste("Residuals for", current_trait), 
       xlab = paste("Residuals ", current_trait))
  qqnorm(y2, main = paste("Residuals for", current_trait))
  qqline(y2, col = "blue")
  
  # fit regression model (apply log transformation)
  obj <- lm(log(pheno[[current_trait]]) ~ pheno$sex + pheno$age)
  y2 <- obj$residuals
  hist(y2, main = paste("Residuals for \n Log transformed", current_trait), 
       xlab = paste("Residuals log", current_trait))
  qqnorm(y2, main = paste("Residuals for \n Log transformed", current_trait))
  qqline(y2, col = "blue")
}
```

LABA10, LABA11, LABA12, LABA13, LABA14: convert percentage to fraction, then apply shrinkage $\epsilon + x(1-2\epsilon)$, with $\epsilon = 0.01$, before a logit transformation.

```{r fig.width=7, fig.height=3.5}
epsilon = 0.01
logit <- function(p) {
  log(p / (1 - p))
}

cleaned_pheno <- pheno %>%
  mutate(
    LABA10 = LABA10 / 100,                       
    shrinkage_logit = logit(epsilon + LABA10 * (1 - 2 * epsilon))
  )

par(mfrow = c(1, 2))
obj <- lm(LABA10 ~ sex + age, data = cleaned_pheno)
y2 <- obj$residuals
hist(y2, main = "Residuals for LABA10", 
     xlab = "Residuals LABA10")
qqnorm(y2, main = "Residuals for LABA10")
qqline(y2, col = "blue")

obj <- lm(shrinkage_logit ~ sex + age, data = cleaned_pheno)
y2 <- obj$residuals
hist(y2, main = "Residuals for logit LABA10", 
     xlab = "Residuals logit LABA10")
qqnorm(y2, main = "Residuals for logit LABA10")
qqline(y2, col = "blue")
```

```{r fig.width=7, fig.height=3.5}
cleaned_pheno <- pheno %>%
  mutate(
    LABA11 = LABA11 / 100,                       
    shrinkage_logit = logit(epsilon + LABA11 * (1 - 2 * epsilon))
  )

par(mfrow = c(1, 2))
obj <- lm(LABA11 ~ sex + age, data = cleaned_pheno)
y2 <- obj$residuals
hist(y2, main = "Residuals for LABA11", 
     xlab = "Residuals LABA11")
qqnorm(y2, main = "Residuals for LABA11")
qqline(y2, col = "blue")

obj <- lm(shrinkage_logit ~ sex + age, data = cleaned_pheno)
y2 <- obj$residuals
hist(y2, main = "Residuals for logit LABA11", 
     xlab = "Residuals logit LABA11")
qqnorm(y2, main = "Residuals for logit LABA11")
qqline(y2, col = "blue")
```

```{r fig.width=7, fig.height=3.5}
cleaned_pheno <- pheno %>%
  mutate(
    LABA12 = LABA12 / 100,                       
    shrinkage_logit = logit(epsilon + LABA12 * (1 - 2 * epsilon))
  )

par(mfrow = c(1, 2))
obj <- lm(LABA12 ~ sex + age, data = cleaned_pheno)
y2 <- obj$residuals
hist(y2, main = "Residuals for LABA12", 
     xlab = "Residuals LABA12")
qqnorm(y2, main = "Residuals for LABA12")
qqline(y2, col = "blue")

obj <- lm(shrinkage_logit ~ sex + age, data = cleaned_pheno)
y2 <- obj$residuals
hist(y2, main = "Residuals for logit LABA12", 
     xlab = "Residuals logit LABA12")
qqnorm(y2, main = "Residuals for logit LABA12")
qqline(y2, col = "blue")
```

```{r fig.width=7, fig.height=3.5}
cleaned_pheno <- pheno %>%
  mutate(
    LABA13 = LABA13 / 100,                       
    shrinkage_logit = logit(epsilon + LABA13 * (1 - 2 * epsilon))
  )

par(mfrow = c(1, 2))
obj <- lm(LABA13 ~ sex + age, data = cleaned_pheno)
y2 <- obj$residuals
hist(y2, main = "Residuals for LABA13", 
     xlab = "Residuals LABA13")
qqnorm(y2, main = "Residuals for LABA13")
qqline(y2, col = "blue")

obj <- lm(shrinkage_logit ~ sex + age, data = cleaned_pheno)
y2 <- obj$residuals
hist(y2, main = "Residuals for logit LABA13", 
     xlab = "Residuals logit LABA13")
qqnorm(y2, main = "Residuals for logit LABA13")
qqline(y2, col = "blue")
```

```{r fig.width=7, fig.height=3.5}
cleaned_pheno <- pheno %>%
  mutate(
    LABA14 = LABA14 / 100,                       
    shrinkage_logit = logit(epsilon + LABA14 * (1 - 2 * epsilon))
  )

par(mfrow = c(1, 2))
obj <- lm(LABA14 ~ sex + age, data = cleaned_pheno)
y2 <- obj$residuals
hist(y2, main = "Residuals for LABA14", 
     xlab = "Residuals LABA14")
qqnorm(y2, main = "Residuals for LABA14")
qqline(y2, col = "blue")

obj <- lm(shrinkage_logit ~ sex + age, data = cleaned_pheno)
y2 <- obj$residuals
hist(y2, main = "Residuals for logit LABA14", 
     xlab = "Residuals logit LABA14")
qqnorm(y2, main = "Residuals for logit LABA14")
qqline(y2, col = "blue")
```

SLPA54 -- log transformation
```{r fig.width=7, fig.height=3.5}
cleaned_pheno <- pheno %>%
  filter(!is.na(SLPA54), SLPA54 > 0) %>%
  filter(!is.na(sex), !is.na(age)) 

par(mfrow = c(1, 2))
obj <- lm(as.numeric(SLPA54) ~ sex + age, data = cleaned_pheno)
y2 <- obj$residuals
hist(y2, main = "Residuals for SLPA54", 
     xlab = "Residuals SLPA54")
qqnorm(y2, main = "Residuals for SLPA54")
qqline(y2, col = "blue")

obj <- lm(log(as.numeric(SLPA54)) ~ sex + age, data = cleaned_pheno)
y2 <- obj$residuals
hist(y2, main = "Residuals for log SLPA54", 
     xlab = "Residuals log SLPA54")
qqnorm(y2, main = "Residuals for log SLPA54")
qqline(y2, col = "blue")
```