```{r}
library(tidyverse)
library(genio)

setwd('/hpc/group/ochoalab/tt207/meta_analysis_aim/hchs-sol')
```

# write id's for sex (not available in fam)
```{r}
pheno = read.csv("/hpc/group/ochoalab/zh105/project2/hchs-sol/Ia/heritability/hchs_phen.csv", header = TRUE)
# read eigen
grm <- read_grm( '/hpc/group/ochoalab/zh105/project2/hchs-sol/Ia/data_qc_std_mor' )
eigenvec <- read_eigenvec( '/hpc/group/ochoalab/tt207/meta_analysis_aim/hchs-sol/data_qc_std_mor' )

pheno <- pheno[ match( grm$fam$id, pheno$SUBJECT_ID ), ]
# now that data is aligned, include all PCs as a single, convenient covariate
pheno$PCs <- eigenvec$eigenvec

table(pheno$GENDERNUM)
# 0 - 6893 individuals
male = pheno %>% filter(GENDERNUM == 0) %>% mutate(famid = SUBJECT_ID, id = SUBJECT_ID) %>% select(famid, id)
# 1 - 4938 individuals
female = pheno %>% filter(GENDERNUM == 1) %>% mutate(famid = SUBJECT_ID, id = SUBJECT_ID) %>% select(famid, id)
# write group id's
write.table(male, "ids_male.txt", col.names = TRUE, 
            row.names = FALSE, quote = FALSE)
write.table(female, "ids_female.txt", col.names = TRUE, 
            row.names = FALSE, quote = FALSE)
```

set 1 of traits
* HEIGHT
* BMI
* WAIST_HIP
* INSULIN_FAST
```{r}
hist(pheno$HEIGHT) 
hist(pheno$BMI)
hist(pheno$WAIST_HIP)
hist(pheno$INSULIN_FAST)
qqnorm(pheno$INSULIN_FAST)
qqline(pheno$INSULIN_FAST, col = "blue")
hist(log(pheno$INSULIN_FAST))
qqnorm(log(pheno$INSULIN_FAST))
qqline(log(pheno$INSULIN_FAST), col = "blue")
```
# set 2 of traits
* INSULIN_FAST -- log transformed
* LABA70: glucose fasting (mg/dL) -- log transformed
* LABA76: glucose post OGTT (mg/dL) -- log transformed
* LABA101: cystatin C (mg/L) -- log transformed
* LABA91: High-sensitivity C-Reactive Protein (mg/L) -- log transformed
* LABA66: Total cholesterol (mg/dL) 
```{r}
hist(pheno$LABA70)
hist(log(pheno$LABA70))
qqnorm(log(pheno$LABA70))
qqline(log(pheno$LABA70), col = "blue")

hist(pheno$LABA76)
hist(log(pheno$LABA76))
qqnorm(log(pheno$LABA76))
qqline(log(pheno$LABA76), col = "blue")

hist(pheno$LABA101)
qqnorm(pheno$LABA101)
hist(log(pheno$LABA101))
qqnorm(log(pheno$LABA101))
qqline(log(pheno$LABA101), col = "blue")

hist(pheno$LABA91)
qqnorm(pheno$LABA91)
hist(log(pheno$LABA91))
qqnorm(log(pheno$LABA91))
qqline(log(pheno$LABA91), col = "blue")

pheno_2 = pheno %>% mutate(LOG_INSULIN_FAST = log(INSULIN_FAST),
                           LOG_LABA70 = log(LABA70),
                           LOG_LABA76 = log(LABA76),
                           LOG_LABA101 = log(LABA101),
                           LOG_LABA91 = log(LABA91))
```
# set 3 of traits
* LABA67: Triglycerides (mg/dL) -- log transformed
* LABA68: HDL-cholesterol (mg/dL) 
* LABA69: LDL-cholesterol (mg/dL) 
* ANTA4: Weight (kg)
```{r}
hist(pheno$LABA67)
qqnorm(pheno$LABA67)
hist(log(pheno$LABA67))
qqnorm(log(pheno$LABA67))
qqline(log(pheno$LABA67), col = "blue")

pheno_3 = pheno %>% mutate(LOG_LABA67 = log(LABA67))
```

# set 4 of traits
* ANTA10A: Waist Girth (cm)
* ANTA10B: Hip Girth (cm) ***
* SBPA5: Average Systolic  
* SBPA6: Average Diastolic  



# write covar

```{r}
#traits = c("HEIGHT", "BMI", "WAIST_HIP", "INSULIN_FAST")
#traits = c("LOG_INSULIN_FAST", "LOG_LABA70", "LOG_LABA76", "LOG_LABA101", "LOG_LABA91", "LABA66")
#traits = c("LOG_LABA67", "LABA68", "LABA69", "ANTA4")
traits = c("ANTA10A", "ANTA10B", "SBPA5", "SBPA6")
setwd('/hpc/group/ochoalab/tt207/meta_analysis_aim/hchs-sol')
for (i in seq_along(traits)) {
  current_trait <- traits[i]
  #current_trait_age <- trait_age[i]
  
  # Create the covar dataframe with the current trait and trait_age
  covar <- pheno %>%
    select(SUBJECT_ID, sex = GENDERNUM, all_of(current_trait), PCs) 
  
  
  write.table(covar, paste0("covar_", current_trait, "_trait.txt"), col.names = TRUE,
            row.names = FALSE, quote = FALSE)
}


```

